---
title: "Intermediate Elective"
date: "2026-05-15"
author: "Garrett Mellinger"
format: pdf
execute:
  message: false
  warning: false
---

# Set up

## Packages

```{r}
#|label: packages

# general use
library(tidyverse)
library(janitor)

# file organization
library(here)

# customizing visualizations
library(RColorBrewer)
```


# Data

```{r}
#| label: data

# reading in data
veg <- read_csv(here("data", "veg.csv"))

metadata <- read_csv(here("data", "vp_veg_metadata.csv"))
```

# Wrangling

```{r}
#| label: wrangling veg

poaceae_species <- c(
  "Bromus_hordeaceus",
  "Festuca_perennis",
  "Stipa_pulchra",
  "Polypogon_monspeliensis",
  "Brachypodium_distachyon",
  "Festuca_myuros",
  "Avena_barbata",
  "Hordeum_marinum",
  "Avena_fatua",
  "Parapholis_incurva",
  "Bromus_diandrus",
  "Distichlis_spicata",
  "Bromus_catharticus",
  "Hordeum_murinum",
  "Hordeum_brachyantherum_brachy.",
  "Cortaderia_selloana",
  "Festuca_bromoides",
  "Cynodon_dactylon",
  "Poa_annua",
  "Festuca_microstachys",
  "Polypogon_viridis",
  "Crypsis_schoenoides",
  "Polypogon_interruptus",
  "Hordeum_brachyantherum",
  "Bromus_madritensis",
  "Elymus_triticoides"
)

grassland <- veg |> 
  # keep only NCOS Mesa grassland transects and only grass species
  filter(
    str_detect(transect_name, "GL"),
    site == "ncos",
    psoc %in% poaceae_species
  )
```



# 1. Explain your inspiration

I chose Cristina Cametti’s “IKEA most popular designers” for my inspiration. This is a reasonable choice for the vegetation dataset because there are a large number of different grass species present on the NCOS mesa, just as there are a large number of different IKEA designers. Each grass species found on the NCOS Mesa will be represented by its own color. The height of each bar will represent the total count of unique plots that species was observed in. The labels at the end of each bar will describe whether the grass is native or not.

# 2. Planning


# 3. Code

```{r}

# TidyTuesday 03/11/2020 - IKEA Furniture

data <- ikea %>%
  group_by(category, designer) %>%
  tally() %>%
  ungroup()


data <- data %>%
  slice(-1:-143) %>%
  filter(!designer == "IKEA of Sweden")


data$designer <- str_replace_all(data$designer, "/IKEA of Sweden", "")
data$designer <- str_replace_all(data$designer, "IKEA of Sweden/", "")


dt <- data %>%
  arrange(desc(n)) %>%
  filter(n>15)


dt <- mutate_at(dt, vars(category), as.factor)
dt <- mutate_at(dt, vars(designer), as.factor)
glimpse(dt)


# Start by setting the plot 
# Set a number of 'empty bar' to add at the end of each group
empty_bar <- 2
# Add lines to the initial dataset
to_add <- matrix(NA, empty_bar, ncol(dt))
colnames(to_add) <- colnames(dt)
dt <- rbind(dt, to_add)
dt$id <- seq(1, nrow(dt))

# Get the name and the y position of each label
label_data <- dt
number_of_bar <- nrow(label_data)
angle <- 90 - 360 * (label_data$id-0.5) /number_of_bar     
label_data$hjust <- ifelse(angle < -90, 1, 0)
label_data$angle <- ifelse(angle < -90, angle+180, angle)

# palette
pal <- brewer.pal(11, "Spectral")


# Plotting
p <- ggplot(dt)+
  geom_bar(aes(x=as.factor(id), y=n, fill=category), stat="identity") +
  scale_fill_manual(values=pal)+
  ylim(-50,90) +
  theme_bw() + 
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    plot.margin = unit(rep(-1,4,), "cm")) +
  annotate(geom = "text",
           x=0,y=-50,
           hjust=.5, vjust=1,
           label="IKEA most popular\n designers",
           size=4.3, lineheight=.8,
           family="Staatliches Regular",
           color="black") +
  annotate(geom = "text",
           x=0,y=200,
           vjust=1,
           hjust=.5,
           label = "Data:  Kaggle & TidyTuesday | Graphic: Cristina Cametti - @CameCry",
           size=2.5,
           color="black") +
  coord_polar(start = 0)  +
  geom_text(data=label_data, aes(x=id, y=n+1, label=designer, hjust=hjust), color="black", 
            fontface="bold",alpha=0.6, size=3.4, angle=label_data$angle, inherit.aes = FALSE ) 

p
```





# 4. Describing my process
In 1-3 sentences each, describe:

your process of using someone else’s code to create a visualization
challenges you encountered as you made your visualization
how your visualization differs from visualizations you made in class, for assignments, or for a previous elective


