---
title: "Intermediate Elective"
date: "2026-05-15"
author: "Garrett Mellinger"
format: pdf
execute:
  message: false
  warning: false
---

# Set up

## Packages

```{r}
#|label: packages

# general use
library(tidyverse)
library(janitor)

# file organization
library(here)

# customizing visualizations
library(RColorBrewer)
```


# Data

```{r}
#| label: data

# reading in data
veg <- read_csv(here("data", "veg.csv"))

metadata <- read_csv(here("data", "vp_veg_metadata.csv"))
```


# 1. Explain your inspiration

I chose Cristina Cametti’s “IKEA most popular designers” for my inspiration. This is a reasonable choice for the vegetation dataset because there are a large number of different grass species present on the NCOS mesa, just as there are a large number of different IKEA designers. Each grass species found on the NCOS Mesa will be represented by its own color. The height of each bar will represent the total count of unique plots that species was observed in. The labels at the end of each bar will describe whether the grass is native or not.

# 2. Planning


# 3. Code

# Wrangling

```{r}
#| label: wrangling veg

# rolling up Hordeum brachyantherum brachyantherum to species level
veg <- veg |>
  mutate(
    psoc = case_when(
      psoc == "Hordeum_brachyantherum_brachy." ~ "Hordeum_brachyantherum",
      TRUE ~ psoc
    )
  )

poaceae_species <- c(
  "Bromus_hordeaceus",
  "Festuca_perennis",
  "Stipa_pulchra",
  "Polypogon_monspeliensis",
  "Brachypodium_distachyon",
  "Festuca_myuros",
  "Avena_barbata",
  "Hordeum_marinum",
  "Avena_fatua",
  "Parapholis_incurva",
  "Bromus_diandrus",
  "Distichlis_spicata",
  "Bromus_catharticus",
  "Hordeum_murinum",
  "Cortaderia_selloana",
  "Festuca_bromoides",
  "Cynodon_dactylon",
  "Poa_annua",
  "Festuca_microstachys",
  "Polypogon_viridis",
  "Crypsis_schoenoides",
  "Polypogon_interruptus",
  "Hordeum_brachyantherum",
  "Bromus_madritensis",
  "Elymus_triticoides"
)

grassland <- veg |> 
      # keep only 2025 NCOS Mesa grassland transects and only grass species
  filter(
    year == 2025,

    str_detect(transect_name, "GL"),
    site == "ncos",
    psoc %in% poaceae_species
  )

# calculating number of observations for each species
grass_counts <- grassland |>
  count(psoc, sort = TRUE, name = "occurrence_count") |> 
  arrange(desc(occurrence_count))
```

# Coding

```{r}
#| label: circular-grass-plot
#| fig-width: 7
#| fig-height: 6

# create plot-ready data from the already-sorted grass species counts
grass_plot <- grass_counts

# create cleaned species names for the legend
grass_plot <- grass_plot |>
  mutate(
    psoc_clean = str_replace_all(psoc, "_", " ")
  )

# set the number of empty bars to create a visual gap in the circle
empty_bar <- 2 

# create empty rows that will become the gap in the circular plot
to_add <- data.frame(
  psoc = rep(NA, empty_bar),
  occurrence_count = rep(NA, empty_bar),
  psoc_clean = rep(NA, empty_bar)
)

# add the empty rows to the plot dataset
grass_plot <- bind_rows(grass_plot, to_add)

# create an id column to control where each bar appears around the circle
grass_plot <- grass_plot |>
  mutate(id = row_number())

# calculate maximum observation count for proportional plot sizing
max_count <- max(grass_plot$occurrence_count, na.rm = TRUE)

# create label data for count labels around the circular plot
label_data <- grass_plot |>
  filter(!is.na(occurrence_count)) |>
  mutate(
    angle = 90 - 360 * (id - 0.5) / nrow(grass_plot),
    hjust = ifelse(angle < -90, 1, 0),
    angle = ifelse(angle < -90, angle + 180, angle)
  )

# create one color for each grass species
palette <- colorRampPalette(brewer.pal(11, "Spectral"))(
  length(unique(na.omit(grass_plot$psoc_clean)))
)

# create the circular bar plot
p <- ggplot(grass_plot) +
  # draw one bar for each grass species
  geom_col(
    aes(
      x = as.factor(id),
      y = occurrence_count,
      fill = psoc_clean
    ),
    width = 1
  ) +
  # apply custom colors and remove NA from legend
  scale_fill_manual(
    values = palette,
    name = "Grass species",
    na.translate = FALSE
  ) +
  # make the plot proportions closer to the original IKEA circular bar chart
  scale_y_continuous(
    limits = c(-max_count * 0.70, max_count * 1.12),
    expand = c(0, 0)
  ) +
  # convert the bar chart into a circular layout
  coord_polar(start = 0, clip = "off") +
  # add count labels around the outside of the circle
  geom_text(
    data = label_data,
    aes(
      x = as.factor(id),
      y = occurrence_count + max_count * 0.06,
      label = occurrence_count,
      hjust = hjust,
      angle = 0,
    ),
    size = 2.4,
    fontface = "bold",
    color = "black",
    inherit.aes = FALSE
  ) +
  # add title inside the center hole, like the original inspiration plot
  annotate(
    geom = "text",
    x = 0,
    y = -max_count * 0.52,
    hjust = 0.5,
    vjust = 0.65,
    label = "2025 Number of\nObservations of\nGrass Species in\nNCOS Grasslands",
    size = 4.3,
    lineheight = 0.85,
    fontface = "bold",
    color = "black"
  ) +
  # use a clean background
  theme_minimal() +
  # remove axes, gridlines, and extra plot clutter
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),

    # remove external title spacing because title is inside the plot
    plot.title = element_blank(),
    plot.subtitle = element_blank(),

    # reduce outer blank space
    plot.margin = margin(0, 0, 0, 0),

    # keep legend on the right but compact
    legend.position = "right",
    legend.direction = "vertical",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.20, "cm"),
    legend.spacing.y = unit(0.01, "cm"),
    legend.box.spacing = unit(0.05, "cm"),
    legend.margin = margin(0, 0, 0, -8),
    legend.box.margin = margin(0, 0, 0, -10),

    # force white background
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  ) +
  # keep legend in one compact vertical column
  guides(
    fill = guide_legend(
      ncol = 1,
      byrow = TRUE,
      keywidth = unit(0.20, "cm"),
      keyheight = unit(0.20, "cm")
    )
  )

# show the finished plot
p
```


```{r}
#| label: saving-plot

# save the final circular grass species plot as a PNG file
ggsave(
  filename = here("code", "grass_species_commonness.png"),
  plot = p,
  width = 6,
  height = 3,
  units = "in",
  dpi = 300,
  bg = "white"
)
```


```{r}

# TidyTuesday 03/11/2020 - IKEA Furniture

data <- ikea %>%
  group_by(category, designer) %>%
  tally() %>%
  ungroup()


data <- data %>%
  slice(-1:-143) %>%
  filter(!designer == "IKEA of Sweden")


data$designer <- str_replace_all(data$designer, "/IKEA of Sweden", "")
data$designer <- str_replace_all(data$designer, "IKEA of Sweden/", "")


dt <- data %>%
  arrange(desc(n)) %>%
  filter(n>15)


dt <- mutate_at(dt, vars(category), as.factor)
dt <- mutate_at(dt, vars(designer), as.factor)
glimpse(dt)


# Start by setting the plot 
# Set a number of 'empty bar' to add at the end of each group
empty_bar <- 2
# Add lines to the initial dataset
to_add <- matrix(NA, empty_bar, ncol(dt))
colnames(to_add) <- colnames(dt)
dt <- rbind(dt, to_add)
dt$id <- seq(1, nrow(dt))

# Get the name and the y position of each label
label_data <- dt
number_of_bar <- nrow(label_data)
angle <- 90 - 360 * (label_data$id-0.5) /number_of_bar     
label_data$hjust <- ifelse(angle < -90, 1, 0)
label_data$angle <- ifelse(angle < -90, angle+180, angle)

# palette
pal <- brewer.pal(11, "Spectral")


# Plotting
p <- ggplot(dt)+
  geom_bar(aes(x=as.factor(id), y=n, fill=category), stat="identity") +
  scale_fill_manual(values=pal)+
  ylim(-50,90) +
  theme_bw() + 
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    plot.margin = unit(rep(-1,4,), "cm")) +
  annotate(geom = "text",
           x=0,y=-50,
           hjust=.5, vjust=1,
           label="IKEA most popular\n designers",
           size=4.3, lineheight=.8,
           family="Staatliches Regular",
           color="black") +
  annotate(geom = "text",
           x=0,y=200,
           vjust=1,
           hjust=.5,
           label = "Data:  Kaggle & TidyTuesday | Graphic: Cristina Cametti - @CameCry",
           size=2.5,
           color="black") +
  coord_polar(start = 0)  +
  geom_text(data=label_data, aes(x=id, y=n+1, label=designer, hjust=hjust), color="black", 
            fontface="bold",alpha=0.6, size=3.4, angle=label_data$angle, inherit.aes = FALSE ) 

p
```





# 4. Describing my process
In 1-3 sentences each, describe:

your process of using someone else’s code to create a visualization
challenges you encountered as you made your visualization
how your visualization differs from visualizations you made in class, for assignments, or for a previous elective


